A sheaf--function dictionary for algebraic tori over local fields

### 2013-09-20 shanghai.tex
Added Warfield to the bibliography.

%\input{introduction}
%In his celebrated thesis, John Tate writes 
%\begin{quotation}
%    {\it Concerning the characters of $k^*$ [the
%    multiplicative group of a $\mathfrak{p}$-adic field], the
%    situation is different from that of $k^+$.  First of all, we are
%    interested in continuous multiplicative maps $\alpha \mapsto
%    c(\alpha)$ of $k^*$ into the complex numbers, not only in the
%    bounded ones, and shall call such a map a quasi-character,
%    reserving the word ``character'' for the conventional character of
%    absolute value $1$. Secondly, we shall find no model for the group
%    of quasi-characters, or even for the group of characters, though
%    such a model would be of the utmost importance.}
%  \cite{tate:67a}*{\S 2.3}
%\end{quotation}
%\noindent
In this paper we find a dictionary between quasicharacters of algebraic tori over local fields and quasicharacter sheaves.

What is a quasicharacter sheaf?
Let $G$ be a smooth, commutative group scheme locally of finite
type over a finite field $\Fq$; $G$ need not be reductive, connected, or of
finite type over $\Fq$. Fix a prime $\ell$ invertible in $\Fq$
and write $m : \GxG \to G$ for the multiplication morphism,
$\bG$ for the base change of $G$ to $\bFq$ and
$\Frob{G} : \bG \to \bG$ for the Frobenius automorphism.
A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank-1 and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms;
$\mu$ and $\phi$ are required to satisfy natural conditions explained below.

@@

While we have not yet fully explored the applications of such a geometrization, we want to highlight
some of the potential benefits of Theorem~\ref{thm:application} here. Characters on $T$ have limited
functoriality: they can be pulled back along maps to $T$, but it is less clear what to do with them
in the presence of maps from $T$ to other groups.  Sheaves, on the other hand, come
equipped with a host of functors, including pushforward. 
We hope to use this to our advantage as we work toward an adaptation of Lusztig's theory of characters sheaves to the study of characters of admissible representations of reductive $p$-adic groups that extends beyond the case of $p$-adic tori.

DeBacker and Reeder have recently applied Deligne-Lusztig representations to part of the
local Langlands correspondence. % \cite{reeder-debacker:09a}.  
After developing Deligne-Lusztig representations, 
the next step Lusztig took in studying representations
of finite groups of Lie type was to introduce character sheaves for reductive groups over $\Fq$.
As suggested above above, we hope that our quasicharacter sheaves 
can provide a stepping stone to a similar theory for
representations of $p$-adic groups.

The scope of the Langlands program has grown explosively in the past several decades, and
different parts of it apply to different kinds of representations.  In particular, the classic local Langlands
correspondence addresses complex (or $\ell$-adic) representations of real and $p$-adic groups, while
the geometric Langlands program aims at representations of groups in characteristic $p$.  The fact that
characteristic $p$ group schemes arise as the space underlying quasicharacter sheaves, regardless of
the characteristic of $K$, provides a potential bridge allowing the methods of the geometric Langlands
program to be applied to the classic context.

Finally, we are tantalized by the possibility of providing an alternate perspective on local class field theory
through quasicharacter sheaves.  The local Langlands correspondence for tori provides an isomorphism
\[
\Hh^1(\Weil{K}, \hat{T}_\ell) \cong \Hom(T(K), \EEx).
\]
A geometric description of the resulting isomorphism $\Hh^1(\Weil{K}, \hat{T}_\ell) \cong \QCiso{\GN{T}}$
would yield a different description of the Artin reciprocity map.

@@

  Now suppose that $[z] \in \Hh^1(\Fq,\Hom(X,\EEx))$ with
  $\rho([z]) = 1$; write $\tilde{f}$ for $z(\Frob{\Fq})$.  The injectivity
  of $\rho$ is equivalent to the existence of a homomorphism $\psi
  \colon X \rightarrow \EEx$ with
  $$\tilde{f}(x) = \frac{\psi(\Frob{\Fq}(x))}{\psi(x)}$$
  for $x \in X$.  In fact, we may replace the construction of $\psi$
  for all $\ZZ[\Gal(\bFq/\Fq)]$-modules $X$ with the construction of
  analogous $\psi$s for a different class of modules.

  Since the action of $\Gal(\bFq/\Fq)$ is continuous, there is an $m$
  so that $\Frob{\Fq}^m$ acts trivially on $X$; write $\Fqm$ for the degree
  $m$ extension of $\Fq$ in $\bFq$ and $\Gamma_m$ for $\Gal(\Fqm/\Fq)$.  Since
  $T^m-1$ factors as a squarefree product of cyclotomic polynomials,
  the Chinese remainder theorem expresses $\ZZ[\Gamma_m]$ as a product
  of $\ZZ[\zeta_d]$ for $d$ dividing $m$.  Each $\ZZ[\zeta_d]$ is the
  maximal order in a number field and thus $\ZZ[\Gamma_m]$ is a
  product of Dedekind domains.  Therefore $X$ decomposes into a direct
  sum of cyclic $\ZZ[\Gamma_m]$-modules \cite{brandal:CommRingsFGModDecompose}, \David{more specific citation} \Clifton{I'm tracking this down.}
  each of the \David{Add a reference to this section}
  form $\ZZ[\Gamma_m] / I$ for some ideal $I \subset \ZZ[\Gamma_m]$.
  The ideal $I$ corresponds to a direct product of ideals $I_d$ in the
  $\ZZ[\zeta_d]$, and within each $\ZZ[\zeta_d]$ the Chinese remainder
  theorem allows a further decomposition of $\ZZ[\zeta_d] / I_d$ as a
  direct sum of $\ZZ[\zeta_d]$-modules of the form $\ZZ[\zeta_d]/P^s$.
  So if we can construct, for every cyclotomic ring $\ZZ[\zeta_d]$,
  prime ideal $P \subset \ZZ[\zeta_d]$ and positive integer $s$, a
  $\psi \colon \ZZ[\zeta_d]/P^s \rightarrow \EEx$ with
  \begin{equation} \label{eq:psi-condition}
    \tilde{f}(x) = \frac{\psi(\zeta_d \cdot x)}{\psi(x)},
  \end{equation}
  then the existence of the original $\psi$ follows because $X$ is a
  direct sum of such $\ZZ[\zeta_d]/P^s$.

>>>>

%\input{introduction}
%In his celebrated thesis, John Tate writes 
%\begin{quotation}
%    {\it Concerning the characters of $k^*$ [the
%    multiplicative group of a $\mathfrak{p}$-adic field], the
%    situation is different from that of $k^+$.  First of all, we are
%    interested in continuous multiplicative maps $\alpha \mapsto
%    c(\alpha)$ of $k^*$ into the complex numbers, not only in the
%    bounded ones, and shall call such a map a quasi-character,
%    reserving the word ``character'' for the conventional character of
%    absolute value $1$. Secondly, we shall find no model for the group
%    of quasi-characters, or even for the group of characters, though
%    such a model would be of the utmost importance.}
%  \cite{tate:67a}*{\S 2.3}
%\end{quotation}
%\noindent
In this paper we find a dictionary between quasicharacters of algebraic tori over local fields and quasicharacter sheaves. 
%
What is a quasicharacter sheaf?
Let $G$ be a smooth, commutative group scheme locally of finite
type over a finite field $\Fq$; $G$ need not be reductive, connected, or of
finite type over $\Fq$. Fix a prime $\ell$ invertible in $\Fq$
and write $m : \GxG \to G$ for the multiplication morphism,
$\bG$ for the base change of $G$ to $\bFq$ and
$\Frob{G} : \bG \to \bG$ for the Frobenius automorphism.
A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank-1 and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms;
$\mu$ and $\phi$ are required to satisfy natural conditions explained below.

@@

While we have not yet fully explored the applications of such a geometrization, we want to highlight
some of the potential benefits of Theorem~\ref{thm:application}. 
%\begin{enumerate}
%\item
Characters on $T$ have limited
functoriality: they can be pulled back along maps to $T$, but it is less clear what to do with them
in the presence of maps from $T$ to other groups.  Sheaves, on the other hand, come
equipped with a host of functors, including pushforward. 
We hope to use this to our advantage as we work toward an adaptation of Lusztig's theory of characters sheaves to the study of characters of admissible representations of reductive $p$-adic groups that extends beyond the case of $p$-adic tori. 
%\item
DeBacker and Reeder have recently applied Deligne-Lusztig representations to part of the
local Langlands correspondence. % \cite{reeder-debacker:09a}.  
After developing Deligne-Lusztig representations, 
the next step Lusztig took in studying representations
of finite groups of Lie type was to introduce character sheaves for reductive groups over $\Fq$.
We hope that our quasicharacter sheaves 
can provide a stepping stone to a similar theory for
representations of $p$-adic groups. 
%\item
The scope of the Langlands program has grown explosively in the past several decades, and
different parts of it apply to different kinds of representations.  In particular, the classic local Langlands
correspondence addresses complex (or $\ell$-adic) representations of real and $p$-adic groups, while
the geometric Langlands program aims at representations of groups in characteristic $p$.  The fact that
characteristic $p$ group schemes arise as the space underlying quasicharacter sheaves, regardless of
the characteristic of $K$, provides a potential bridge allowing the methods of the geometric Langlands
program to be applied to the classic context. 
%\item
Finally, we are tantalized by the possibility of providing an alternate perspective on local class field theory
through quasicharacter sheaves.  The local Langlands correspondence for tori provides an isomorphism
\[
\Hh^1(\Weil{K}, \hat{T}_\ell) \cong \Hom(T(K), \EEx).
\]
A geometric description of the resulting isomorphism $\Hh^1(\Weil{K}, \hat{T}_\ell) \cong \QCiso{\GN{T}}$
would yield a different description of the Artin reciprocity map.
%\end{enumerate}

@@

  Now suppose that $[z] \in \Hh^1(\Fq,\Hom(X,\EEx))$ with
  $\rho([z]) = 1$; write $\tilde{f}$ for $z(\Frob{\Fq})$.  The injectivity
  of $\rho$ is equivalent to the existence of a homomorphism $\psi
  \colon X \rightarrow \EEx$ with
  $$\tilde{f}(x) = \frac{\psi(\Frob{\Fq}(x))}{\psi(x)}$$
  for $x \in X$.  In fact, we may replace the construction of $\psi$
    \Clifton{I'm not sure I understand what you mean here.}
  for all $\ZZ[\Gal(\bFq/\Fq)]$-modules $X$ with the construction of
  analogous $\psi$s for a different class of modules.

  The $\ZZ[\Gal(\bFq/\Fq)]$-module $X$ decomposes into 
  a finite direct sum of cyclic $\ZZ[\Gal(\bFq/\Fq)]$-modules. 
  To see this, observe first that since the action of $\Gal(\bFq/\Fq)$ is continuous, 
  there is a positive integer $m$ so that $\Frob{\Fq}^m$ acts trivially on $X$; 
  write $\Fqm$ for the degree $m$ extension of $\Fq$ in $\bFq$ 
  and $\Gamma_m$ for $\Gal(\Fqm/\Fq)$.  Since
  $T^m-1$ factors as a squarefree product of cyclotomic polynomials,
  the Chinese remainder theorem expresses $\ZZ[\Gamma_m]$ as a product
  of $\ZZ[\zeta_d]$ for $d$ dividing $m$.  Each $\ZZ[\zeta_d]$ is the
  maximal order in a number field and thus $\ZZ[\Gamma_m]$ is a
  product of Dedekind domains. Combining \cite{Warfield:Decomposability}*{Theorem 3}
  and \cite{brandal:CommRingsFGModDecompose}*{Lemma 4.1}, it now follows that 
  $X$ decomposes into a direct sum of cyclic $\ZZ[\Gamma_m]$-modules, 
  each of the form $\ZZ[\Gamma_m] / I$ for some ideal $I \subset \ZZ[\Gamma_m]$.
  The ideal $I$ corresponds to a direct product of ideals $I_d$ in the
  $\ZZ[\zeta_d]$, and within each $\ZZ[\zeta_d]$ the Chinese remainder
  theorem allows a further decomposition of $\ZZ[\zeta_d] / I_d$ as a
  direct sum of $\ZZ[\zeta_d]$-modules of the form $\ZZ[\zeta_d]/P^s$.
  So if we can construct, for every cyclotomic ring $\ZZ[\zeta_d]$,
  prime ideal $P \subset \ZZ[\zeta_d]$ and positive integer $s$, a
  $\psi \colon \ZZ[\zeta_d]/P^s \rightarrow \EEx$ with
  \begin{equation} \label{eq:psi-condition}
    \tilde{f}(x) = \frac{\psi(\zeta_d \cdot x)}{\psi(x)},
  \end{equation}
  then the existence of the original $\psi$ follows because $X$ is a
  direct sum of such $\ZZ[\zeta_d]/P^s$.
